What We Eval in the Shadows

### 2021-04-04 paper/main.tex
Update pipeline figure
Pass on section 3 and 4

\begin{abstract}
  \noindent Most dynamic languages allow users to turn text into code
  using various functions, often named \eval, with language-dependent
  semantics. The widespread use of these reflective functions hinders
  static analysis and prevents compilers from performing
  optimizations. This paper aims to provide a better sense of why
  programmers use \eval. Understanding why \eval is used in practice
  is key to finding ways to mitigate its negative impact. We have
  reasons to believe that reflective feature usage is language and
  application domain specific; we focus on data science code written
  in R, and compare our results to previous work that analyzed web
  programming in JavaScript. Dynamic analysis of a corpus of
  \CranAllPackages libraries and \KaggleCode end-user code confirms that
  \eval is indeed in widespread use; R's \eval is more dangerous in
  some ways, and safer in others, than what was previously reported
  for JavaScript.
  %% TODO ::: Add a sentence about summarizing why users resort to
  %% eval -- Jan
\end{abstract}

@@

\vspace{2mm}\noindent\emph{The What and How.} One significant benefit of
choosing R is that every package in the CRAN repository is curated and comes
with examples of typical usage. This gives us a large code base that we can
analyze dynamically. To observe \eval we built a two-level monitoring
infrastructure:\footnote{Our infrastructure is open source and publicly
available, our anonymized code is at {\tt http://url.com}, a more complete
artifact will be submitted to the artifact evaluation committee. } we can
monitor R programs by instrumentation --- this gives us access to many
user-visible properties of R programs --- but we can also monitor the
inner-workings of the R interpreter --- this allows us to capture details not
exposed at the source level. Dynamic analysis is limited, it can only observe
behaviors triggered by the particular inputs passed to a program. Luckily, CRAN
libraries come with many tests and use-cases. The choice of corpus is crucial.
Our corpus has been constructed to reflect the levels of sophistication of the R
community. We distinguish between \emph{CRAN packages} (\CranAllPackages curated
packages that pass stringent quality checks and are equipped with tests and
sample data) and \emph{Kaggle scripts} (\KaggleUnique end-user written programs
that performs a particular data analysis task). It is reasonable to expect that
\eval usage differ between these datasets: the libraries represent a lively
ecosystem with new libraries added each day, while end user code is often thrown
together, run once, and never revisited.

@@

\mypara{Base.} The base dataset contains 13 core libraries performing basic
arithmetics, statistics, and operating system related functionalities. These
libraries are bundled with R and are executed pervasively.

\mypara{CRAN.} The Comprehensive R Archive
Network\footnote{\url{cran.r-project.org}} is the largest curated repository of
R packages. It hosts \CranAvailablePackagesRnd packages with 6 new one submitted
each day~\cite{Ligges2017}. Packages accepted to CRAN are authored by
experienced developers and abide to a number of well-formedness rules
automatically checked on each commit. Furthermore, each package comes with
sample data. There are three sources of runnable code in a package: \emph{tests,
examples} and \emph{vignettes} -- respectively, unit tests, code snippets from
the documentation, and long-form use-cases written in RMarkdown.
Examples and vignettes can be turned into scripts by extracting the relevant
code. We obtained \CranPackages that call \eval. These packages represent
\CranPackagesRatio of all packages or \CranCodeRatio of the entire code size of
CRAN. From these we extracted \CranRunnableScripts scripts that contain
\CranRunnableCode lines of code (\CranRunnableCodeExamplesRnd in examples,
\CranRunnableCodeVignettesRnd in vignettes and \CranRunnableCodeTestsRnd in
tests).

\mypara{Scripts.} Kaggle is an online platform for data
science.\footnote{\url{kaggle.com}} It allows users to submit problems and
compete to solve them. Solutions, called \emph{kernels}, are uploaded as plain
scripts or notebooks. The quality of code on Kaggle is not uniform. Each kernel
is runnable, and thus there is one script per kernel. Input data is provided. We
obtained \KaggleKernels kernels. Since Kaggle is not curated, used SHA-1 hashes
to identify and remove \KaggleDuplicates duplicate entries. The remaining
\KaggleUnique kernels are unique solutions to \KaggleCompetitions competition.
They contain \KaggleCode lines of R code. Only \KaggleWithEvals kernels call
\eval.

@@

\begin{figure}[!h]\hspace{-5mm}
  \includegraphics[width=1.05\linewidth]{pipeline.pdf}
  \caption{Pipeline UPDATE!!}\label{fig:pipeline}
\end{figure}

\medskip
\begin{compactenum}
\item \emph{Download.} Packages are downloaded from CRAN. For scripts, a web
  crawler retrieves code and the Kaggle command line tool gets data.
  Installation is complicated by native dependencies which are not properly
  documented and thus hard to automatically resolve.
\item \emph{Extract.} Given installed packages and scripts, the next step is to
  create runnable programs. The \genthat tool helpfully extracts all runnable
  code snippets from a package and turns each of these into a self-standing
  program~\cite{issta18}. Some Kaggle kernels are already scripts, for those
  nothing more needs be done. Others kernels are packages as notebooks, either
  as Rmarkdown or Jupyter, for those we use \c{knitr} to extract runnable code.
  The body of each extracted program is instrumented with calls our dynamic
  analyzer to ensure that we only record calls to \eval from the code of
  interest and not from bootstrapping or execution harness operations.
\item \emph{Trace.} Each program is executed using our dynamic analysis tool
  which is heavily instrumented interpreter that captures calls to \eval and
  many other fine-grained runtime events. Packages are run twice, once to
  capture \eval calls originating from package code, and a second time to
  capture calls coming from the base libraries. To avoid any interference, each
  program is run in its own process.
\item \emph{Analyze.} Finally, analysis output is merged, cleaned and summarized
  in a post-processing phase driven by series of R scripts. The summarized data
  is then analyzed in RMarkdown notebooks to gather insights. All figures and
  numbers appearing in the paper are generated automatically. Figures are
  produced in PDF by \c{ggplot2}, numbers are exported as \LaTeX macros.
\end{compactenum}

\medskip\noindent
The code extraction and tracing steps of the pipeline are run in
parallel~\cite{GNUparallel}. Servers have identical environment thanks to docker
images with all dependencies installed. The pipeline consists of 3.5K lines of
Make, R and shell scripts.

\subsection{Analysis}

The dynamic analysis is performed by \rdyntrace, a modified R virtual machine
based on GNU R 4.0.2 that exposes low-level callbacks for a variety of runtime
events~\cite{oopsla19a}. As the GNU R bytecode compiler is written in R, we turn
it off to avoid recording \eval in its code, and fallback on the interpreter.
This study requires recording a large number of events, calls to \eval,
side-effects, function entries and exits, environments, etc. One challenge we
encountered during the analysis was that R does not produce source references
for top-level blocks. That is to say, a method whose body would \c{eval(x)}
would not have debug information that we can use to identify that particular
\eval. The same expression surrounded by braces would have a line number and
source reference. This is unfortunately not easily fixed in the R
implementation. Instead, we extended the dynamic analysis tool to add missing
source references heuristically.

The analysis has some limitation. Even with the extension described above, there
are still \CranUndefinedRnd \eval calls without source references. This occurs
when a reference to the \eval function is passed as argument to a higher-order
functions or when the \eval call originates from native code. However, these
missing source references account for a meager \CranUndefinedRatio of all calls;
they are unlikely to affect our results. Other limitations are that we ignore
calls to the native \eval function and to the alternate \c{rlang::tidy\_eval}
function which uses native \eval internally. The \c{rlang} package introduces a
new kind of promise called \c{quosure} which is evaluated by \c{tidy\_eval}.
None of these limitation should invalidate our conclusions.

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Usage Metrics}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

This section focuses on CRAN packages and reports statistics about the usage of
\eval. Calls to \eval appear in {\bf 28\%} of the packages hosted on CRAN. There
are \Corpus packages with an \eval, all programs extracted from these packages
were executed, any run that did not exercise one of the \eval call sites in the
packages was discarded. This left \Nbruns runs of some of the \Corpus packages
to generate the data presented here. There were \Allcalls invocations of \eval
and its variants occurring in \Triggeredpkgs packages.\footnote{Not all packages
with static call sites of \eval were observed triggering that call site, this
can be chalked down to incomplete code coverage and occasional infrastructure
failures of our analysis infrastructure.} Figure~\ref{freq} summarizes the
frequency of dynamic calls to \eval with the left column being the number of
calls and the right, the number of packages in that range. There are \Fewcalls
packages with low \eval frequency, fewer than 100 calls; and a similar number of
packages, \Manycalls to be precise, that use \eval more than 1,000 times. One
package, \Maxcallspack, calls it \Maxcalls times and thus accounts for over half
of the observed calls.

@@

\begin{wrapfigure}{r}{5.8cm}
 \includegraphics[width=0.5\textwidth, trim=4.8cm 0 6cm 0, clip]{size_loaded_distribution}
\caption{Loaded code} \label{fig:sizedistribution}
\end{wrapfigure}

To estimate how much executable code is injected through \eval, we measure the
the number of abstract syntax tree nodes as returned by the parser; for example,
\c{x+1} counts as 3.\footnote{We attempted to measure string lengths after
de-parsing expressions, but these measurements were dominated by de-parsed data
objects, which could range in the MBs.} The median argument size is
\Medianszeval (due to the many symbols) and the average is \Avgszeval nodes. The
largest \eval input observed is \Maxszeval\,-- a significant chunk of code.
Figure~\ref{fig:sizedistribution} shows the distribution of sizes for arguments
of fewer or equal to 25 nodes. The x-axis is the size of arguments in number of
nodes, and the y-axis is the count of arguments with that size. The size drops
rapidly, with few observations larger than 15 nodes. The long tail is omitted
from the graph for legibility.

@@

\begin{figure}[h!]
\begin{tabular}{@{}c@{}c@{}}
\begin{minipage}{7.5cm}
 \includegraphics[width=\textwidth]{events_per_pack_small}
\end{minipage}&\begin{minipage}{7.5cm}
  \includegraphics[width=\textwidth]{events_per_pack_large}
\end{minipage}\\[-3mm]
\small (a) Small & \small (b) Large
\end{tabular}
 \caption{Instructions per \eval} \label{ev}
\end{figure}

>>>>

\begin{abstract}
  \noindent Most dynamic languages allow users to turn text into code
  using various functions, often named \eval, with language-dependent
  semantics. The widespread use of these reflective functions hinders
  static analysis and prevents compilers from performing
  optimizations. This paper aims to provide a better sense of why
  programmers use \eval. Understanding why \eval is used in practice
  is key to finding ways to mitigate its negative impact. We have
  reasons to believe that reflective feature usage is language and
  application domain specific; we focus on data science code written
  in R, and compare our results to previous work that analyzed web
  programming in JavaScript. Dynamic analysis of a corpus of
  \CranPackages libraries and \KaggleCode end-user code confirms that
  \eval is indeed in widespread use; R's \eval is more dangerous in
  some ways, and safer in others, than what was previously reported
  for JavaScript.
  %% TODO ::: Add a sentence about summarizing why users resort to
  %% eval -- Jan
\end{abstract}

@@

\vspace{2mm}\noindent\emph{The What and How.} One significant benefit of
choosing R is that every package in the CRAN repository is curated and comes
with examples of typical usage. This gives us a large code base that we can
analyze dynamically. To observe \eval we built a two-level monitoring
infrastructure:\footnote{Our infrastructure is open source and publicly
available, our anonymized code is at {\tt http://url.com}, a more complete
artifact will be submitted to the artifact evaluation committee. } we can
monitor R programs by instrumentation --- this gives us access to many
user-visible properties of R programs --- but we can also monitor the
inner-workings of the R interpreter --- this allows us to capture details not
exposed at the source level. Dynamic analysis is limited, it can only observe
behaviors triggered by the particular inputs passed to a program. Luckily, CRAN
libraries come with many tests and use-cases. The choice of corpus is crucial.
Our corpus has been constructed to reflect the levels of sophistication of the R
community. We distinguish between \emph{CRAN packages} (\CranPackages curated
packages that pass stringent quality checks and are equipped with tests and
sample data) and \emph{Kaggle scripts} (\KaggleUnique end-user written programs
that performs a particular data analysis task). It is reasonable to expect that
\eval usage differ between these datasets: the libraries represent a lively
ecosystem with new libraries added each day, while end user code is often thrown
together, run once, and never revisited.

@@

\mypara{Base.} The base dataset contains \BasePackages core libraries
performing basic arithmetics, statistics, and operating system related
functionalities. These libraries are bundled with R and are executed
pervasively. Together they contain \BaseEvalCallSites calls to \eval in
\BaseFunsWithEvals functions. Some of these functions provide basic R
functionalities such as package loading and thus there is hardly any R code
which does not invoke one of these \eval call sites.

\mypara{CRAN.} The Comprehensive R Archive
Network\footnote{\url{cran.r-project.org}} is the largest curated repository of
R packages. It hosts \CranAvailablePackagesRnd packages with 6 new one submitted
each day~\cite{Ligges2017}. Packages accepted to CRAN are authored by
experienced developers and abide to a number of well-formedness rules
automatically checked on each commit. Furthermore, each package comes with
sample data. There are three sources of runnable code in a package: \emph{tests,
examples} and \emph{vignettes} -- respectively, unit tests, code snippets from
the documentation, and long-form use-cases written in RMarkdown.
Examples and vignettes can be turned into scripts by extracting the relevant
code. From \CranPackages packages we extracted \CranRunnableScripts scripts that contain
\CranRunnableCode lines of code (\CranRunnableCodeExamplesRnd in examples,
\CranRunnableCodeVignettesRnd in vignettes and \CranRunnableCodeTestsRnd in
tests).

\mypara{Kaggle.} The Kaggle website\footnote{\url{kaggle.com}} is an online
platform for data science. It allows users to submit problems and compete to
solve them. Solutions, called \emph{kernels}, are uploaded as plain scripts or
notebooks. The quality of code on Kaggle is not uniform. Each kernel is
runnable, and thus there is one script per kernel. Input data is provided. We
obtained \KaggleKernels kernels. Since Kaggle is not curated, used SHA-1 hashes
to identify and remove \KaggleDuplicates duplicate entries. The remaining
\KaggleUnique kernels are unique solutions to \KaggleCompetitions competition.
They contain \KaggleCode lines of R code. Only \KaggleWithEvals kernels call
\eval.

@@

\begin{figure}[!h]\hspace{-5mm}
  \includegraphics[width=1.05\linewidth]{pipeline.pdf}
  \caption{Pipeline}\label{fig:pipeline}
\end{figure}

\medskip
\begin{compactenum}
\item \emph{Download.} Packages are downloaded from CRAN. For scripts, a web
  crawler retrieves code and the Kaggle command line tool gets data.
  Installation is complicated by native dependencies which are not properly
  documented and thus hard to automatically resolve.
\item \emph{Extract.} Given installed packages and scripts, the next step is to
  create runnable programs. The \genthat tool helpfully extracts all runnable
  code snippets from a package and turns each of these into a self-standing
  program~\cite{issta18}. Some Kaggle kernels are already scripts, for those
  nothing more needs be done. Others kernels are packages as notebooks, either
  as Rmarkdown or Jupyter, for those we use \c{knitr} to extract runnable code.
  The body of each extracted program is instrumented with calls our dynamic
  analyzer to ensure that we only record calls to \eval from the code of
  interest and not from bootstrapping or execution harness operations.
\item \emph{Trace.} Each program is executed using our dynamic analysis tool
  which is heavily instrumented interpreter that captures calls to \eval and
  many other fine-grained runtime events. Packages are run twice, once to
  capture \eval calls originating from package code, and a second time to
  capture calls coming from the base libraries. To avoid any interference, each
  program is run in its own process.
  As the GNU R bytecode compiler is written in R, we turn it off
  to avoid recording \eval in its code, and fallback on the interpreter.
\item \emph{Analyze.} Finally, analysis output is merged, cleaned and summarized
  in a post-processing phase driven by series of R scripts. The summarized data
  is then analyzed in RMarkdown notebooks to gather insights. All figures and
  numbers appearing in the paper are generated automatically. Figures are
  produced in PDF by \c{ggplot2}, numbers are exported as \LaTeX macros.
\end{compactenum}

\medskip\noindent The code extraction and tracing steps of the pipeline are run
in parallel~\cite{GNUparallel} orchestrated by a Makefile. Servers have
identical environment thanks to docker images with all dependencies installed.

\subsection{Dynamic Analysis}

The dynamic analysis is performed by \rdyntrace, a modified R virtual machine
based on GNU R 4.0.2 that exposes low-level callbacks for a variety of runtime
events~\cite{oopsla19b}.

\mypara{Tracer.} The tracer registers callbacks to all \eval functions as well
as a few other ones that allow us to better identify where do the expressions
come from. For example, we taint the results of calls to \c{parse} (turns
string into expression) and \c{match.call} (reflects on current call). We
also capture calls to R API for dynamic code loading (\eg\xspace \c{library},
\c{require} and \c{source}). Next, we subscribe to the events related to a
variable definition and assignment allowing us to record side effects that
happen in environments while evaluating code in \eval. One challenge we
encountered during the analysis was that essentially R only provides source
references for top-level calls in a function body or in the block surrounded by
braces. That is to say, a method whose body would \c{eval(x)} would not have
debug information that we can use to identify that particular \eval. The same
expression surrounded by braces would have a line number and source reference.
This is unfortunately not easily fixed in the R implementation. Instead, we
extended the dynamic analysis tool to attach synthetic source code references
to all \eval call sites by traversing AST.  The tracer is implemented as an R
package in 3.2K C++ and 1.3K of R code. For performance reasons, most of the
tracing is done in C++. While in theory the implementation is rather
straightforward, what makes it hard in practice is that (1) some of the core R
functionality such as package loading is implemented using \eval which then
leaks to the tracer in unexpected places, (2) lazy evaluation makes it harder
to analyze function arguments while tracing (3) running large amount of
real-world code triggers exercises a lot of corner cases of highly
underspecified R behavior.

\mypara{Limitations.} The analysis has some limitation. Even with the extension
described above, there are still \PkgUndefinedRnd \eval calls without source
references. This occurs when a reference to the \eval function is passed as
argument to a higher-order functions or when the \eval call originates from
native code. However, these missing source references account for a meager
\PkgUndefinedRatio of all calls; they are unlikely to affect our results. Other
limitations are that we ignore calls to the native \eval function and to the
alternate \c{rlang::tidy\_eval} function which uses native \eval internally.
The \c{rlang} package introduces a new kind of promise called \c{quosure} which
is evaluated by \c{tidy\_eval}. None of these limitation should invalidate our
conclusions.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Usage Metrics}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{wrapfigure}{r}{5cm} \hspace*{-12mm}
  \includegraphics[width=75mm]{pkgs-eval-callsites-hist.pdf} \caption{CRAN
  \eval call sites}%
  \label{fig:pkgs-eval-callsites-hist}
\end{wrapfigure}
%
This section focuses on CRAN packages and reports statistics about the usage of
\eval. There are \PkgEvalCallSites calls to \eval in \PkgPackages packages.
These represent \PkgPackagesRatio of all packages or \PkgCodeRatio of the
entire code size of CRAN. Over half of the packages have less than 3 \eval
call sites and except \MaxEvalCallSitesPackage (\MaxEvalCallSitesCount \eval
calls), all packages contain less than \MaxEvalCallSitesRest \eval call sites
(\cf Fig~\ref{fig:pkgs-eval-callsites-hist}) These \eval calls appear in
\PkgFunsWithEval R functions (\CranFunsWithEvalRatio of all R functions in
CRAN).

For the dynamic analysis, we execute all the \CranRunnableScripts programs
extracted from \CranPackages packages. Any run that did not exercise one of the
\eval call sites in the packages was discarded. This left \Nbruns runs of some
of the \Corpus packages to generate the data presented here. There were
\Allcalls invocations of \eval and its variants occurring in \Triggeredpkgs
packages. Out of \PkgEvalCallSitesRnd \eval call sites, \PkgHitEvalCallSites
were hit at least once\footnote{Not all packages with static call sites of
\eval were observed triggering that call site, this can be chalked down to
incomplete code coverage and occasional tracer failure (\PkgFailedProgramsRatio
of programs do not terminate successfully or within 30 minute timeout).}. On
average, the analysis covers \PkgHitEvalCallSitesAvgRatio of package \eval call
sites, which similar to the \PkgCodeCoverage of average package code coverage.
Figure~\ref{freq} summarizes the frequency of dynamic calls to \eval with the
left column being the number of calls and the right, the number of packages in
that range. There are \Fewcalls packages with low \eval frequency, fewer than
100 calls; and a similar number of packages, \Manycalls to be precise, that use
\eval more than 1,000 times.  One package, \Maxcallspack, calls it \Maxcalls
times and thus accounts for over half of the observed calls.

@@

To estimate how much executable code is injected through \eval, we measure the
the number of abstract syntax tree nodes as returned by the parser; for example,
\c{x+1} counts as 3.\footnote{We attempted to measure string lengths after
de-parsing expressions, but these measurements were dominated by de-parsed data
objects, which could range in the MBs.} The median argument size is
\Medianszeval (due to the many symbols) and the average is \Avgszeval nodes. The
largest \eval input observed is \Maxszeval\,-- a significant chunk of code.
Figure~\ref{fig:sizedistribution} shows the distribution of sizes for arguments
of fewer or equal to 25 nodes.
%
\begin{wrapfigure}{r}{5.8cm}
 \hspace*{-10mm}
 \includegraphics[width=0.5\textwidth, trim=4.8cm 0 6cm 0, clip]{size_loaded_distribution}
 \caption{Loaded code} \label{fig:sizedistribution}
\end{wrapfigure}
%
The x-axis is the size of arguments in number of
nodes, and the y-axis is the count of arguments with that size. The size drops
rapidly, with few observations larger than 15 nodes. The long tail is omitted
from the graph for legibility.

@@

\begin{figure}[tb!]
\begin{tabular}{@{}c@{}c@{}}
\begin{minipage}{7.5cm}
 \includegraphics[width=\textwidth]{events_per_pack_small}
\end{minipage}&\begin{minipage}{7.5cm}
  \includegraphics[width=\textwidth]{events_per_pack_large}
\end{minipage}\\[-3mm]
\small (a) Small & \small (b) Large
\end{tabular}
 \caption{Instructions per \eval} \label{ev}
\end{figure}
